Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-04-13 manuscript/manuscript_file.Rmd

@@

![Location of sampling sites.  Color of the points indicates whether the waterbody was functioning as a source or sink of nitrous oxide (N~2~O).  Point size indicates the dissolved N~2~O concentration.](manuscript_figures/figure1.png)

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production.  This finding  runs counter to the long standing paradigm that nitrogen rich surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lentic waterbodies can function as an N~2~O sink.  @Webb_etal_2019 found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months and @Soued_etal_2016 report that 40% of rivers, lakes, and ponds sampled during the summer in a boreal region of Quebec, Canada were functioning as N~2~O sinks.  While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months.  This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

```{r table1, echo = FALSE, cache=TRUE}
rbind(national_stats %>% 
        select(WSA9, estimate, LCL, UCL, post_med_sat, LCI_sat, UCI_sat, post_med_en2o, LCL_en2o, UCL_en2o),
      ecoreg_stats %>% 
        select(WSA9, estimate, LCL, UCL, post_med_sat, LCI_sat, UCI_sat, post_med_en2o, LCL_en2o, UCL_en2o)) %>%
  mutate('N2O undersaturation extent' = paste0(estimate*100, " (", LCL*100, "-", UCL*100, ")"),
         'N2O saturation ratio' = paste0(post_med_sat, " (", LCI_sat, "-", UCI_sat, ")"),
         'N2O emission rate' = paste0(post_med_en2o, " (", LCL_en2o, "-", UCL_en2o, ")")) %>% 
  select(WSA9, 'N2O undersaturation extent', 'N2O saturation ratio', 'N2O emission rate') %>%
  rename('spatial domain' = WSA9) %>%
knitr::kable(caption = "Mean and 95% confidence interval of N2O undersaturation extent, saturation ratio, and emission rate at the national and ecoregional scales")

@@

![Distribution of N~2~O saturation values for the population of waterbodies in the nine major ecoregions.  Vertical black dashed and solid lines indicated median and mean, respectively.  Vertical red line indicates a saturation ratio of 1.](manuscript_figures/figure2.tiff)

#### Controls on N~2~O saturation

N~2~O saturation was related to waterbody size in several ecoregions, but the direction of the relationship varied.  For example, N~2~O saturation increased with waterbody size in the Northern Plains, Upper Midwest, Northern Appalachians, and Western Mountain ecoregions, but decreased with increasing size in the Coastal and Southern Plains (Figure 3).  A pattern common to many of the ecoregions, regardless of the direction of the relationship between N~2~O saturation and waterbody size, is that N~2~O saturation deviates from equilibrium most strongly in the smallest waterbody size class.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as delta N~2~O, decreases linearly with increasing waterbody size class.

NEED TO ADD THE POOLED FIGURE

>>>>

## NOTES
* Table and figure captions should specify if displaying sample vs population estimates.

@@

![Figure 1. Location of sampling sites.  Color of the points indicates whether the waterbody was functioning as a source or sink of nitrous oxide (N~2~O).  Point size indicates the dissolved N~2~O concentration.](manuscript_figures/figure1.png)

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production.  This finding  runs counter to the long standing paradigm that nitrogen rich surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lentic waterbodies can function as an N~2~O sink.  Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months and Soued et al [@Soued_etal_2016] report that 40% of rivers, lakes, and ponds sampled during the summer in a boreal region of Quebec, Canada were functioning as N~2~O sinks **[29% of lakes in Finland (kortelainen et al 2020)  but kortelainen and Soued are mostly pristine lakes]**.  While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months.  This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

```{r table1, echo = FALSE, cache=TRUE}
rbind(national_stats %>% 
        select(WSA9, estimate, LCL, UCL, post_med_sat, LCI_sat, UCI_sat, post_med_en2o, LCL_en2o, UCL_en2o),
      ecoreg_stats %>% 
        select(WSA9, estimate, LCL, UCL, post_med_sat, LCI_sat, UCI_sat, post_med_en2o, LCL_en2o, UCL_en2o)) %>%
  mutate('N2O undersaturation extent' = paste0(estimate*100, " (", LCL*100, "-", UCL*100, ")"),
         'N2O saturation ratio' = paste0(post_med_sat, " (", LCI_sat, "-", UCI_sat, ")"),
         'N2O emission rate' = paste0(post_med_en2o, " (", LCL_en2o, "-", UCL_en2o, ")")) %>% 
  select(WSA9, 'N2O undersaturation extent', 'N2O saturation ratio', 'N2O emission rate') %>%
  rename('spatial domain' = WSA9) %>%
knitr::kable(caption = "Table 1. Mean and 95% confidence interval of N2O undersaturation extent, saturation ratio, and emission rate at the national and ecoregional scales")

@@

![Figure 2. Distribution of N~2~O saturation values for the population of waterbodies in the nine major ecoregions.  Each blue line (n=500) represents the distribution of values from a single realization of the posterior prediction.  Vertical black dashed and solid lines indicated median and mean, respectively.  Vertical red line indicates a saturation ratio of 1.](manuscript_figures/figure2.tiff)

#### Controls on N~2~O saturation
N~2~O saturation was related to waterbody size in several ecoregions, but the direction of the relationship varied.  For example, N~2~O saturation increased with waterbody size in the Northern Plains, Upper Midwest, Northern Appalachians, and Western Mountain ecoregions, but decreased with increasing size in the Coastal and Southern Plains (Figure 3A).  The inconsistent relationship between N~2~O saturation and waterbody size may be attributable differences in nitrate concentrations across ecoregions.  The model identified an interaction between waterbody surface area and nitrate concentration (expressed categorically) where N~2~O concentration decreases with increasing lake size at high NO3 concentrations, but increases, or remains constant at low nitrate concentrations. THIS IS BECAUSE THERE CAN'T BE MUCH N2O BIOGEOCHEMISTRY WITHOUT INORGANIC N LAYING AROUND.

![Figure 3. Relationship between delta N~2~O and lake size category across A) the population of CONUS waterbodies and B) within each of the nine major CONUS ecoregions.  Red arrows in panel A illustrate the magnitude of N~2~O disequilibrium in the smallest lake size class.  Data are inferences made to the population of CONUS waterbodies.](manuscript_figures/figure3.tiff)

![Figure 4. Partial dependence plot illustrating the relationship between N~2~O concentration and waterbody surface area as a function of nitrate concentration.](manuscript_figures/N2O vs surface area x NO3 partial dependence plot.png)

Regardless of the direction of the relationship between N~2~O saturation and waterbody size, N~2~O saturation deviates from equilibrium most strongly in the smallest waterbody size class.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases linearly with increasing waterbody size class (Figure 3B).  Waterbody size can affect dissolved gas concentration through several mechanisms.  In lakes with low volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column.  Since water volume to benthic surface area ratios tend to scale positively with lake surface area, benthic N~2~O biogeochemistry will have a greater affect on dissolved N~2~O concentration in small lakes, regardless of whether the sediments function as a net sink or source of N~2~O **[see Wetzel for references]**.  Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes **[need a citation for N2O ~ temperature relationship]**.  Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (Beaulieu et al N2O Harsha Lake and USACE N2O papers).

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequlibrim in smaller waterbodies.  Air-water gas exchange will vent excess N~2~O from surface waters or replenish depleted N~2~O in undersaturated waterbodies by facilitating the invasion of atmospheric N~2~O.  Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium.  One of the primary drivers of air-water gas exchange in lentic waters is wind speed (Vachon and Prairie reference).  Wind speeds to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation.  Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished.

**Not sure where I am going with this paragraph.  Maybe scratch it.  Mostly included as a placeholder for these references**
While previous surveys have documented relationships between lake size and dissolved methane concentration (Pighini et al. 2018, Juutinen et al 2009, Borges et al. 2011), another greenhouse gas, to the best of our knowledge, this is the first study to document a relationship between lake size and $\Delta$ N~2~O.  Perhaps because previous surveys have included little variation in waterbody size (Webb et al. 2019, Kortelainen et al) or water quality (Soued et al. 2015).
